Adaptive evolution with aneuploidy and mutation

### 2022-01-12 ms/ms.tex
remove tau from model, just mention in results

%%%%%%%% Fig illustation
\begin{figure}[b!]
  %\centering
  \begin{subfigure}[t]{0.4\textwidth}
      \caption{
        \textbf{A. Single-locus model}
      }
      \centering
      \includegraphics[height=1.8in]{../figures/Fig1-A.pdf}      
      \label{fig:model1}
  \end{subfigure}
  %\\
  \begin{subfigure}[t]{0.4\textwidth}
  	  \caption{
        \textbf{B. Multi-locus model}
      }
      \centering
      \includegraphics[height=1.8in]{../figures/Fig1-B.pdf}      
      \label{fig:model2}
  \end{subfigure}
  \caption{
    \textbf{Model illustrations.}
    \textbf{(A)} In the single-locus model, the four genotypes are: euploid wild-type, \euwt; euploid mutant, \eumt; aneuploid wild-type, \anwt; and aneuploid mutant, \anmt.
    Overall there are two possible trajectories from \euwt\; to \eumt.
    \textbf{(B)} In the multi-locus model, each genotype is characterized by its karyotype, \euwt\; or \anwt, and the number of accumulated beneficial mutations, denoted by stars. 
    In both panels arrows denote transitions between genotypes, with transitions rates: $\mu$ and $\tau \mu$, beneficial mutation rates in euploid and aneuploid cells; $\delta$, aneuploidy rate.
    % TODO IK: change k mu to tau mu in figure; also k mu is too close to the arrow right now; change alpha to delta in the figure 
  }
  \label{fig:models}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%

@@

Transitions between the genotypes occur as follows~(\autoref{fig:models}A): Beneficial mutations from \euwt\; to \eumt\; occur with probability $\mu$, the mutation rate, and from \anwt\; to \anmt\; with probability $\tau \mu$, where $\tau$ is the fold-change in the rate of beneficial mutations in aneuploid cells. By default, we assume $\tau=33/32$, as the wild-type \yeast strains in the experiments by \citep{Yona2012} are diploid with 32 chromosomes, and the aneuploid strains are trisomic, with 33 chromosomes. We neglect back-mutations (i.e., from \eumt\; to \euwt\; and from \anmt\; to \anwt).
Aneuploidy is formed by chromosome mis-segregation, so that cells transition from \euwt\; to \anwt\; and from \anmt\ to \eumt\ with probability $\delta$, the aneuploidy rate. That is, we assume chromosomes are gained and lost at the same rate, and we neglect events that form a less-fit genotype (i.e., \anwt\; to \euwt\; and \eumt\; to \anmt).
The fitness values of the four genotypes are given by \autoref{table:single-locus}.

@@

The initial population has $N$ cells with genotype \euwt.
The effect of natural selection on the frequency $f_i$ of genotype $i = 2n, 2n+1, 2n+1^*, \text{or } 2n^*$ is given by
    \begin{equation} \label{eq:selection-single} 
      f^s_i = \frac{f_i w_i}{\bar{w}} \;,
    \end{equation}
where the fitness values $w_i$ are given in \autoref{table:single-locus} and $\bar{w} = \sum_{j}{f_j w_j}$ is the population mean fitness.
The effect of mutation and aneuploidy on genotype frequencies is given by
    \begin{equation} \label{eq:mutation-aneuploidy-single}
    \begin{aligned}
      &f^m_{2n} &=&\; (1 - \delta - \mu) f^s_{2n}  \;,\\
      &f^m_{2n+1} &=&\; \delta f^s_{2n} + (1 - \tau \mu) f^s_{2n+1}  \;,\\
      &f^m_{2n+1^*} &=&\; \tau \mu f^s_{2n+1} + (1-\delta) f^s_{2n+1^*}  \;,\\
      &f^m_{2n^*} &=&\; \mu f^s_{2n} + \delta f^s_{2n+1} + f^s_{2n^*}  \;.
    \end{aligned}
    \end{equation}
Finally, random genetic drift is modeled using a multinomial distribution~\citep{Otto2007},
    \begin{equation} \label{eq:drift-single}
      \vec{f'} \sim \frac{1}{N} \cdot \mathit{Mult}(N,\ \vec{f^m}) \;,
    \end{equation}
where $\vec{f^m}=(f^m_{2n}\ ,\ f^m_{2n+1}\ ,\ f^m_{2n+1^*}\ ,\ f^m_{2n^*})$ are the frequencies of the genotypes after mutation and aneuploidy, $\vec{f'}$ are the genotype frequencies in the next generation, and $Mult(N,\ \vec{f})$ is a multinomial distribution parameterized by the population size $N$ and the genotype frequencies $\vec{f}$.
Overall, the change in genotype frequencies from one generation to the next is given by the transformation $f_i \to f'_i$.

@@

A genotype is characterized by its karyotype, \euwt\; or \anwt, and the number of accumulated beneficial mutations, which can be zero or more.
The selection coefficient of the $i$-th accumulated mutation in each individual, $s_i$, is drawn from an exponential distribution with expected value $s$: $s_i \sim \mathit{Exp{(s)}}$.
The rest of the parameters ($N$, $\mu$, $\tau$, $\delta$, $b$, $c$) are the same as in the single-locus model. However, since the multi-locus model allows several mutations with smaller fitness effects to accumulate, we expect the mutation rate to be higher compared to the single-locus model, which focuses on a single, large-effect mutation.
% TODO : add tau to model fixed at 33/32.

@@

The effects of mutation and aneuploidy on genotype frequencies is more elaborate than in the single-locus model.
Genotype $i$ is classified according to its karyotype (\euwt\ or \anwt), the number of accumulated beneficial mutations ($k \ge 0$), and their fitness ($w_i$).
Each offspring cell inherits these properties from its mother cell.
Then, with probability $\mu$ or $\tau \mu$ for euploid and aneuploid cells,
respectively, a new beneficial mutation is accumulated, such that the number of mutations is $k+1$, and its effect $s_{k+1}$ is drawn from an exponential distribution with expected value $s$, such that the contribution of the mutations to the fitness is $\prod_{j=0}^{k+1}{(1+s_j)}$.
Next, euploid offspring become aneuploid with probability $\delta$, and aneuploid offspring become euploid with probability $\delta_L$.

@@

\paragraph{Likelihood function.} 
Because our model, just like the Wright-Fisher model, is non-linear and stochastic, computing the distribution of fixation time $T(g)$ of genotype $g$ for use in the likelihood function is intractable (it is even hard to use a diffusion-equation approximation due to the model having multiple genotypes, rather than just two).
We overcome this problem by approximating the likelihood using simulations. We simulate 1,000 experiments per parameter vector $\theta = (\mu, \tau, \delta, s, b, c)$, resulting in a set of simulated observations $\tilde{\vec X} = \{\tilde{X}_i\}_{i=1}^{1000}$. We then compute the approximate likelihood,
\begin{equation}\begin{aligned}
\label{eq:heatstress-likelihood}
\likelihood(\theta) = &\ P^4(200 \le T(2n+1) \le 450) \cdot 
	\Big[1 - \\
	&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)- \\
	&	P_{\tilde{\vec X}}^4\big(!\{1700 < T(2n^*) < 2350\} \mid 200 \le T(2n+1) \le 450\big)+ \\
	&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \land !\{1700 < T(2n^*) < 2350\} \mid 200 \le T(2n+1) \le 450\big) 
	\Big]\;,
\end{aligned}\end{equation}
where $!\{\ldots\}$ is the "logical not" operator, $P^4(\ldots)$ is the 4th power of $P(\ldots)$, and all probabilities $P_{\tilde{\vec X}}(\ldots)$ are approximated from the results of the simulations $\tilde{\vec X}$. For example, $P_{\tilde{\vec X}}\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)$ is approximated by taking simulations in which \anwt\; fixed before generation 450 but not before generation 200, and computing the fraction of such simulations in which \eumt\; did not fix by generation 1,700, and hence aneuploidy did not extinct before generation 1,700.

@@

We also tried to inference parameters of the single-locus model with different $\tau$: 1, 2, 5, 10 and 100. It was seen that for 1, 2, 5 the posterior was not significantly different from the model with $\tau=33/32$, while for $\tau=10$ the estimated $\mu\ $ was a little lower, $1.654\substack{+0.879 \\ -1.775}\times10^{-6}$, and for $\tau=100$ the estimated $\mu\ $ was already significantly lower and $\delta\ $ had higher variability. The MAP and 50\% HDI for $\tau=100$: %TODO add to MAP-and-hdi.ipynb and check the values here
$\mu=5.538\substack{+1.611 \\ -5.311}\times10^{-7}$,
$\delta=8.292\substack{+8.988 \\ -4.96}\times10^{-4}$,
$w_{2n+1}=1.023\substack{+0.001 \\ -0.002}$,
$w_{2n+1^*}=1.024\substack{+0.001 \\ -0.001}$,
$w_{2n^*}=1.028\substack{+0.001 \\ -0.002}$.
%TODO: show joint distributions comparison plot (20-diffk.ipynb). WAIC?

% TODO use tau=1 as the default; we also check tau>1 - 33/32, 2, 5, 10, 100.
% TODO support figure of joint posterior delta vs mu per tau=1, 33/32, 2, 5, 10

@@

\pagebreak
% Acknowledgements
{\small
\section*{Acknowledgements}
We thank Yitzhak Pilpel, Orna Dahan, Lilach Hadany, Judith Berman, David Gresham, Shay Covo, Martin Kupiec, and Tal Simon for discussions and comments.
This work was supported in part by 
the Israel Science Foundation (YR 552/19) and
Minerva Stiftung Center for Lab Evolution (YR).
}

>>>>

%%%%%%%% Fig illustation
\begin{figure}[b!]
  %\centering
  \begin{subfigure}[t]{0.4\textwidth}
      \caption{
        \textbf{A. Single-locus model}
      }
      \centering
      \includegraphics[height=1.8in]{../figures/Fig1-A.pdf}      
      \label{fig:model1}
  \end{subfigure}
  %\\
  \begin{subfigure}[t]{0.4\textwidth}
  	  \caption{
        \textbf{B. Multi-locus model}
      }
      \centering
      \includegraphics[height=1.8in]{../figures/Fig1-B.pdf}      
      \label{fig:model2}
  \end{subfigure}
  \caption{
    \textbf{Model illustrations.}
    \textbf{(A)} In the single-locus model, the four genotypes are: euploid wild-type, \euwt; euploid mutant, \eumt; aneuploid wild-type, \anwt; and aneuploid mutant, \anmt.
    Overall there are two possible trajectories from \euwt\; to \eumt.
    \textbf{(B)} In the multi-locus model, each genotype is characterized by its karyotype, \euwt\; or \anwt, and the number of accumulated beneficial mutations, denoted by stars. 
    In both panels arrows denote transitions between genotypes, with transitions rates: $\mu$, beneficial mutation rate; $\delta$, aneuploidy rate.
    % TODO IK: remove k (not tau either!); also k mu is too close to the arrow right now; change alpha to delta in the figure 
  }
  \label{fig:models}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%

@@

Transitions between the genotypes occur as follows~(\autoref{fig:models}A): Beneficial mutations from \euwt\; to \eumt\; and from \anwt\; to \anmt\; occur with probability $\mu$, the mutation rate. We neglect back-mutations (i.e., from \eumt\; to \euwt\; and from \anmt\; to \anwt).
Aneuploidy is formed by chromosome mis-segregation, so that cells transition from \euwt\; to \anwt\; and from \anmt\ to \eumt\ with probability $\delta$, the aneuploidy rate. That is, we assume chromosomes are gained and lost at the same rate, and we neglect events that form a less-fit genotype (i.e., \anwt\; to \euwt\; and \eumt\; to \anmt).
The fitness values of the four genotypes are given by \autoref{table:single-locus}.

@@

The initial population has $N$ cells with genotype \euwt.
The effect of natural selection on the frequency $f_i$ of genotype $i = 2n, 2n+1, 2n+1^*, \text{or } 2n^*$ is given by
    \begin{equation} \label{eq:selection-single} 
      f^s_i = \frac{f_i w_i}{\bar{w}} \;,
    \end{equation}
where the fitness values $w_i$ are given in \autoref{table:single-locus} and $\bar{w} = \sum_{j}{f_j w_j}$ is the population mean fitness.
The effect of mutation and aneuploidy on genotype frequencies is given by
    \begin{equation} \label{eq:mutation-aneuploidy-single}
    \begin{aligned}
      &f^m_{2n} &=&\; (1 - \delta - \mu) f^s_{2n}  \;,\\
      &f^m_{2n+1} &=&\; \delta f^s_{2n} + (1 - \mu) f^s_{2n+1}  \;,\\
      &f^m_{2n+1^*} &=&\; \mu f^s_{2n+1} + (1-\delta) f^s_{2n+1^*}  \;,\\
      &f^m_{2n^*} &=&\; \mu f^s_{2n} + \delta f^s_{2n+1} + f^s_{2n^*}  \;.
    \end{aligned}
    \end{equation}
Finally, random genetic drift is modeled using a multinomial distribution~\citep{Otto2007},
    \begin{equation} \label{eq:drift-single}
      \vec{f'} \sim \frac{1}{N} \cdot \mathit{Mult}(N,\ \vec{f^m}) \;,
    \end{equation}
where $\vec{f^m}=(f^m_{2n}\ ,\ f^m_{2n+1}\ ,\ f^m_{2n+1^*}\ ,\ f^m_{2n^*})$ are the frequencies of the genotypes after mutation and aneuploidy, $\vec{f'}$ are the genotype frequencies in the next generation, and $Mult(N,\ \vec{f})$ is a multinomial distribution parameterized by the population size $N$ and the genotype frequencies $\vec{f}$.
Overall, the change in genotype frequencies from one generation to the next is given by the transformation $f_i \to f'_i$.

@@

A genotype is characterized by its karyotype, \euwt\; or \anwt, and the number of accumulated beneficial mutations, which can be zero or more.
The selection coefficient of the $i$-th accumulated mutation in each individual, $s_i$, is drawn from an exponential distribution with expected value $s$: $s_i \sim \mathit{Exp{(s)}}$.
The rest of the parameters ($N$, $\mu$, $\delta$, $b$, $c$) are the same as in the single-locus model. However, since the multi-locus model allows several mutations with smaller fitness effects to accumulate, we expect the mutation rate to be higher compared to the single-locus model, which focuses on a single, large-effect mutation.

@@

The effects of mutation and aneuploidy on genotype frequencies is more elaborate than in the single-locus model.
Genotype $i$ is classified according to its karyotype (\euwt\ or \anwt), the number of accumulated beneficial mutations ($k \ge 0$), and their fitness ($w_i$).
Each offspring cell inherits these properties from its mother cell.
Then, a new beneficial mutation is accumulated with probability $\mu$, such that the number of mutations is $k+1$, and its effect $s_{k+1}$ is drawn from an exponential distribution with expected value $s$, such that the contribution of the mutations to the fitness is $\prod_{j=0}^{k+1}{(1+s_j)}$.
Next, euploid offspring become aneuploid with probability $\delta$, and aneuploid offspring become euploid with probability $\delta_L$.

@@

\paragraph{Likelihood function.} 
Because our model, just like the Wright-Fisher model, is non-linear and stochastic, computing the distribution of fixation time $T(g)$ of genotype $g$ for use in the likelihood function is intractable (it is even hard to use a diffusion-equation approximation due to the model having multiple genotypes, rather than just two).
We overcome this problem by approximating the likelihood using simulations. We simulate 1,000 experiments per parameter vector $\theta = (\mu, \delta, s, b, c)$, resulting in a set of simulated observations $\tilde{\vec X} = \{\tilde{X}_i\}_{i=1}^{1000}$. We then compute the approximate likelihood,
\begin{equation}\begin{aligned}
\label{eq:heatstress-likelihood}
\likelihood(\theta) = &\ P^4(200 \le T(2n+1) \le 450) \cdot 
	\Big[1 - \\
	&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)- \\
	&	P_{\tilde{\vec X}}^4\big(!\{1700 < T(2n^*) < 2350\} \mid 200 \le T(2n+1) \le 450\big)+ \\
	&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \land !\{1700 < T(2n^*) < 2350\} \mid 200 \le T(2n+1) \le 450\big) 
	\Big]\;,
\end{aligned}\end{equation}
where $!\{\ldots\}$ is the "logical not" operator, $P^4(\ldots)$ is the 4th power of $P(\ldots)$, and all probabilities $P_{\tilde{\vec X}}(\ldots)$ are approximated from the results of the simulations $\tilde{\vec X}$. For example, $P_{\tilde{\vec X}}\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)$ is approximated by taking simulations in which \anwt\; fixed before generation 450 but not before generation 200, and computing the fraction of such simulations in which \eumt\; did not fix by generation 1,700, and hence aneuploidy did not extinct before generation 1,700.

@@

We also tried to infer parameters of the single-locus model when the mutation rate in aneuploid cells, compared to euploid cells, is increased by a factor $\tau$=1, 33/32 (due to an additional chromosome), 2, 5, 10 or 100 (due to genetic instability).
We found that for $\tau=$1, 33/32, 2, and 5 the posterior was not significantly different from the model with $\tau=1$, while for $\tau=10$ the estimated mutation rate was slightly lower, $\mu=1.654\substack{+0.879 \\ -1.775}\times10^{-6}$, and for $\tau=100$ the estimated mutation rate was significantly lower and the aneuploidy rate had higher variability. The MAP and 50\% HDI with $\tau=100$ were: %
TODO add to MAP-and-hdi.ipynb and check the values here
$\mu=5.538\substack{+1.611 \\ -5.311}\times10^{-7}$,
$\delta=8.292\substack{+8.988 \\ -4.96}\times10^{-4}$,
$w_{2n+1}=1.023\substack{+0.001 \\ -0.002}$,
$w_{2n+1^*}=1.024\substack{+0.001 \\ -0.001}$,
$w_{2n^*}=1.028\substack{+0.001 \\ -0.002}$.
We thus continued our analysis under the assumption that $\tau=1$, that is, that the mutation rate in euploid and aneuploid cells is the same.
%TODO: show joint distributions comparison plot (20-diffk.ipynb). WAIC?
% TODO support figure of joint posterior delta vs mu per tau=1, 33/32, 2, 5, 10

@@

\pagebreak
% Acknowledgements
{\small
\section*{Acknowledgements}
We thank Yitzhak Pilpel, Orna Dahan, Lilach Hadany, Judith Berman, David Gresham, Shay Covo, Martin Kupiec, and Tal Simon for discussions and comments.
This work was supported in part by 
the Israel Science Foundation (YR 552/19) and
Minerva Stiftung Center for Lab Evolution (YR).
% TODO add Minerva for Martin; funding for Avihu?
}
